Promise and peril: Agnostic identification methods\\ for detecting differential item functioning

### 2020-03-14 paper/paper.tex
printed second verison!

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={New and old ways to chase a rabbit: Differential item functioning detection methods},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{amsmath}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\usepackage{float}
\usepackage[autostyle, english = american]{csquotes}
\setlength{\parindent}{4em}
\setlength{\parskip}{2em}
\usepackage{setspace}\doublespacing
\AtBeginEnvironment{tabular}{\singlespacing}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{New and old ways to chase a rabbit:\\
Differential item functioning detection methods}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Ben Stenhaug\\
Stanford University}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle
\begin{abstract}
summarize propose extensions insist as dif as student property and on communicating uncertainty
\end{abstract}

@@

\hypertarget{to-do}{%
\section{TO DO}\label{to-do}}

\begin{itemize}
\tightlist
\item
  be sure all my directions match. i think i switched everything to be item easiness at some point but not sure if i did it throughout
\end{itemize}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Inspired by Camilli (\protect\hyperlink{ref-camilli1992conceptual}{1992}), we think differential item functioning (DIF) as bias against a group of students that manifests when an item response model with too few ability dimensions is imposed. From this perspective, the term \enquote{differential item functioning} is, perhaps, a misnomer as DIF is a property of the student, not the item. For example, Ackerman (\protect\hyperlink{ref-ackerman1992didactic}{1992}) describes a scenario in which a test intends to measure a student's math ability, but performance also depends on their verbal ability. In this case, math ability is the \enquote{target ability,} and verbal ability is the \enquote{nuisance ability.} Fitting a unidimensional item response model to this test results in students with low verbal ability receiving a score systematically lower than their true math ability; therein lies DIF.

Contrarily, the usual setup of DIF simulation studies frames DIF as a property of the item. For example, Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) simulate students as belonging to either a reference or focal group. They fix the item easinesses for the reference group, \({b_j}^{\text{ref}}\), to values from a previous study. They set item easinesses for the focal group to \({b_j}^{\text{foc}} = {b_j}^{\text{ref}}\) for items without DIF and \({b_j}^{\text{foc}} = {b_j}^{\text{ref}} + 0.6\) for items with DIF, where 0.6 is the magnitude of DIF in logits. They simulate student ability \({\theta_i}^{\text{ref}} \sim N(0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(-1,1)\) for students in the focal group. They generate item responses according to the Rasch model, which specifies that the probability of student \(i\) responding correctly to item \(j\) is \[P(y_{ij} = 1 | \theta_i, b_j) = \sigma(\theta_i + b_j)\] where \(\sigma(x) = \frac{e^x}{e^x + 1}\) is the standard logistic function.

We find it informative to describe simulation conditions in terms of multidimensional abilities as opposed to item parameters that vary across groups. For example, to translate the Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) simulation from the DIF-as-item-property view to the DIF-as-student-property view, item easiness is set to what was previously \({b_j}^{\text{ref}}\) for all students. Student ability is expanded to two dimensions, the target ability dimension and nuisance ability dimension. The target ability is what was previously just ability where \({\theta_i}^{\text{ref}} \sim N(0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(-1,1)\) for students in the focal group. The nuisance ability is set to \({\eta_i}^{\text{ref}} = 0\) for the reference group and \({\eta_i}^{\text{ref}} = -1\) for the focal group. We now need to use a 2PL model where the slope on target ability \(a_{1j} = 1\) for all items (consistent with the Rasch model) and the slope on nuisance ability \(a_{2j} = 0.6\) for all items with DIF, and \(a_{2j} = 0\) otherwise. Again, 0.6 is the magnitude of DIF in logits. According to the two-dimensional 2PL model, the probability student \(i\) responding correctly to item \(j\) is
\begin{align}
    P(y_{ij} = 1 | \theta_i, \eta_i, a_{1j}, a_{2j}, b_j) = \sigma(a_{1j}\theta_i + a_{2j}\eta_i + b_j).
\end{align}
This translation between views makes explicit that nearly all DIF simulation studies have, perhaps suboptimally, examined the unrealistic scenario in which there is no variation in nuisance ability for students in the same group.

@@

\hypertarget{dif-methods}{%
\section{DIF methods}\label{dif-methods}}

Psychometricians have long been in search of the perfect DIF detection method. In this section, we summarize existing DIF methods and propose extensions of those methods. We limit our discussion to IRT-based methods. As a result, we avoid methods like the Mantel-Haenszel procedure (Holland and Thayer \protect\hyperlink{ref-holland1986differential}{1986}), which muddies the waters by moving away from the IRT framework, and has been shown to perform no better --- and in some cases worse --- than IRT-based methods (Swaminathan and Rogers \protect\hyperlink{ref-swaminathan1990detecting}{1990}). For simplicity, we focus on the Rasch model to isolate the fundamental issues in DIF detection. And, in search of a coherent framework, we sometimes edit names of existing methods. We recognize that others have done the same (e.g.~Kopf, Zeileis, and Strobl \protect\hyperlink{ref-kopf2015framework}{2015}) and that we risk contributing to the proliferation of names.

We use a simple, one-run simulation to demonstrate the methods: 10,000 reference group students and 10,000 focal group students taking an eight-item test. Target ability is simulated \({\theta_i}^{\text{ref}} \sim N(0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(-1,1)\) for students in the focal group. Nuisance ability is set to \({\eta_i}^{\text{ref}} = 0\) and \({\eta_i}^{\text{ref}} = -1\). The slope on target ability is set to \(a_{1j} = 1\) for all items, and the slope on nuisance ability is set to \(a_{2j} = 0.5\) for the last three items (the items with DIF) and \(a_{2j} = 0\) otherwise. From the DIF-as-item-property view, there is no nuisance ability. Instead, \({b_j}^{\text{foc}} = {b_j}^{\text{ref}}\) for the first five items and \({b_j}^{\text{foc}} = {b_j}^{\text{ref}} - 0.5\) for the last three items.

In general, we denote the mean ability of the reference and focal group as \(\mu^\text{ref}\) and \(\mu^\text{foc}\), respectively. We follow the common and inconsequential practice of identifying the scale by setting \(\mu^\text{ref} = 0\). The fundamental challenge for a DIF method is to first get an estimate for \(\mu^\text{foc}\), \(\hat\mu^\text{foc}\). How can group ability be disentangled from the possibility that items are biased? The most common approach is to identify a group of anchor items that are assumed to be DIF-free. These items are used to get \(\hat\mu^\text{foc}\).

\hypertarget{anchor-items}{%
\subsection{Anchor items}\label{anchor-items}}

@@

Meade and Wright (\protect\hyperlink{ref-meade2012solving}{2012}) compared the most commonly used methods and unequivocally recommended the all-others-as-anchors (AOAA). AOAA tests each item for DIF one at a time using all of the other items as anchors. For example, when testing the first item for DIF, all of the other items are used as anchors. This is done using a likelihood ratio test that compares the baseline model, where all item parameters are fixed across groups, to the flexible model, where the parameters of the tested item are freed across groups (Thissen, Steinberg, and Wainer \protect\hyperlink{ref-thissen1993detection}{1993}). Then, when testing the second item for DIF, once again all of the other items (including the first item) are used as anchors, and so on. The items for which the flexible model outperforms the baseline model (typically based on a chi-squared test) are identified as having DIF and the rest of the items become anchor items.

Edelen et al. (\protect\hyperlink{ref-edelen2006identification}{2006}) used AOAA to look for DIF between the English and Spanish versions of the 21-item Mini-Mental State Examination and found that 10 of the 21 items exhibited DIF. How can we be sure its those 10 items and not the other 11 items with DIF? We can't be. Implicit in the use of AOAA is the assumption that all items not being tested do not exhibit DIF, which is, of course, impossible. More practically, it is thought that AOAA will perform well if a small minority of items have DIF or the DIF is balanced such that some items are biased against the focal group and others are biased against the reference group. Undesirably, most applications of DIF methods and many simulation studies do not make explicit the assumptions of the DIF method (Strobl et al. \protect\hyperlink{ref-strobl2018anchor}{2018}). In this way, psychometricians might benefit from adopting economists' habit of explicitly stating assumptions and debating their plausibility.

@@

One way to attempt to counter artificial DIF is with purification through iterative anchor selection. For example, Drasgow (\protect\hyperlink{ref-drasgow1987study}{1987}) started with AOAA, removed items displaying DIF from the anchor set, then repeated the process iteratively --- with items that have been removed from the anchor set allowed to have free parameters across groups in both the baseline and flexible model --- until no more items tested positively. Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) named this technique Iterative-backward-AOAA with \enquote{backward} (as in reverse, not incorrect) referring to beginning with the assumption that all items are DIF-free. We find it clearer to refer to this method as all-others-as-anchors-all-significiant (AOAA-AS). Appending all-significant indicates that anchor selection is done iteratively with all items that show statistical significance for DIF being removed from the anchor set.

AOAA-AS might seem like an improvement but it doesn't solve a fundamental problem of AOAA: What do we do when all items test positive for DIF? With a sufficient sample size and at least one item with DIF, this will necessarily be the case. In our thought experiment, we get the same result with AOAA-AS as we did with AOAA: All items test positive for DIF and there are no anchor items. Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) encountered precisely this problem in their simulation study and chose to select a single anchor item randomly. Woods (\protect\hyperlink{ref-woods2009empirical}{2009}) suggested a more straightforward, one-step method which uses AOAA and selects the, say, four items that exhibit the least amount of DIF. It's unclear if one should proceed if those four items display DIF themselves.

@@

We propose an extension of these methods, all-others-as-anchors-one-at-a-time (AOAA-OAT), which, to our knowledge (and surprise), has not previously been proposed. Like AOAA-AS, AOAA-OAT starts with AOAA, but only the single item exhibiting the most DIF, based on the \(\chi^2\) test statistic, is removed from the anchor set. The process continues iteratively until no new items display DIF. AOAA-OAT and AOAA-AS are similar in that they are both iterative; the difference is that AOAA-OAT takes the more conservative approach of removing only one item in each iteration as opposed to all items that test positive for DIF. As a result, AOAA-OAT is less likely to arrive at the wrong conclusion because of artifical DIF.

@@

Table \ref{table:themethods} summarizes the three all-others-as-anchors methods. It's useful to remember that AOAA is not an iterative procedure, while the methods with a hyphen, AOAA-OAT and AAOA-AS, are iterative procedures with AOAA-OAT being our new, more conservative method.

\begin{table}[H]
\caption{EDIT THIS}
\centering
\begin{tabular}{|p{2.5cm}|p{6cm}|p{4cm}|}
\toprule

@@

{[}ADD IN SIMULATION RESULTS{]}

@@

Bechger and Maris (\protect\hyperlink{ref-bechger2015statistical}{2015}) proposed choosing anchor items by identifying clusters of items that function similarly and then choosing one of those clusters to be the \enquote{anchor cluster.} They pointed out that one way around the unidentifiability issue is to consider relative item parameters. For each group, the relative easinesses for each pair of items can be stored in the matrix \(\mathbf{R}^{\text{ref}}\) with entries \({R_{xy}}^{\text{ref}} = {b_x}^{\text{ref}} - {b_y}^{\text{ref}}\). The ultimate matrix of interest is \(\Delta \mathbf{R} \equiv \mathbf{R}^{\text{ref}}-\mathbf{R}^{\text{foc}}\) which is the \enquote{differences between groups in the pairwise differences in (easiness) between items} {[}p.~323{]}.

The general idea of identifying clusters of items is intriguing. However, their approach is needlessly complicated and they did not describe a process for moving from \(\Delta \mathbf{R}\) to a set of anchor items. Pohl, Stets, and Carstensen (\protect\hyperlink{ref-pohl2017cluster}{2017}) extended their work by proposing one such process. \(\Delta \mathbf{R}\) is skew-symmetric and of rank 1, which means that all information is contained in a single row or column. Accordingly, they use k-means clustering on just the first column of \(\Delta \mathbf{R}\) where the number of clusters, k, is chosen by minimizing BIC. They suggest using a combination of cluster size, cluster homogeneity, and cluster parameter precision to choose which of the clusters is the anchor cluster. Unfortunately, in their simulation study, they find that BIC identifies only a single cluster, so they end up using all of the items as anchors.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Promise and peril: Agnostic identification methods for detecting differential item functioning},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{amsmath}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\usepackage{float}
\usepackage[autostyle, english = american]{csquotes}
\setlength{\parindent}{4em}
\setlength{\parskip}{2em}
\usepackage{setspace}\doublespacing
\AtBeginEnvironment{tabular}{\singlespacing}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{Promise and peril: Agnostic identification methods\\
for detecting differential item functioning}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Ben Stenhaug, Ben Domingue, and Mike Frank\\
Stanford University}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle
\begin{abstract}
It is well known that likelihood ratio tests (LRT) are effective at detecting differential item functioning (DIF) in item response models. However, to use an LRT, the model needs to be identified so that differences in group ability can be disentangled from potential DIF. We summarize existing assumptionless identification (AI) methods and propose a variety of new methods. We conduct a simulation study --- which we believe to be more realistic than most DIF simulation studies in the literature --- and find that one of the proposed new AI methods, All-others-as-anchors-one-at-a-time (AOAA-OAT), significantly outperforms current methods. We also offer a new method, the equal means, multiple imputation logit graph (EM-MILG), that presents clearly all information about possible DIF, including sampling variability in item parameters, to the analyst. \clearpage
\end{abstract}

@@

\hypertarget{intro}{%
\section{Introduction}\label{intro}}

Inspired by Camilli (\protect\hyperlink{ref-camilli1992conceptual}{1992}), we think of differential item functioning (DIF) as a varying relationship between ability and probability of correct response for students from different groups that manifests when an item response model with too few ability dimensions is imposed. From this perspective, the term \enquote{differential item functioning} is, perhaps, a misnomer as DIF is better thought of as a property of the student, as opposed to the item. For example, Ackerman (\protect\hyperlink{ref-ackerman1992didactic}{1992}) describes a scenario in which a test intends to measure a student's math ability, but performance also depends on their verbal ability. In this case, math ability is the \enquote{target ability,} and verbal ability is the \enquote{nuisance ability.} Fitting a unidimensional item response model to this test results in students with low verbal ability receiving a score systematically lower than their true math ability; therein lies DIF.

Contrarily, the usual setup of DIF simulation studies frames DIF as a property of the item. For example, Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) simulate students as belonging to either a reference or focal group. They fix the item easinesses for the reference group, \({b_j}^{\text{ref}}\), to values from a previous study. They set item easinesses for the focal group to \({b_j}^{\text{foc}} = {b_j}^{\text{ref}}\) for items without DIF and \({b_j}^{\text{foc}} = {b_j}^{\text{ref}} - 0.6\) for items with DIF, where 0.6 is the magnitude of DIF in logits. They simulate student ability \({\theta_i}^{\text{ref}} \sim N(0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(-1,1)\) for students in the focal group. They generate item responses according to the Rasch model, which specifies that the probability of student \(i\) responding correctly to item \(j\) is \[P(y_{ij} = 1 | \theta_i, b_j) = \sigma(\theta_i + b_j)\] where \(\sigma(x) = \frac{e^x}{e^x + 1}\) is the standard logistic function.

We find it informative to describe simulation conditions in terms of multidimensional abilities as opposed to item parameters that vary across groups. For example, to translate the Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) simulation from the DIF-as-item-property view to the DIF-as-student-property view, item easiness is set to what was previously \({b_j}^{\text{ref}}\) for all students. Student ability is expanded to two dimensions, the target ability dimension and nuisance ability dimension. The target ability is what was previously just ability where \({\theta_i}^{\text{ref}} \sim N(0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(-1,1)\) for students in the focal group. The nuisance ability is set to \({\eta_i}^{\text{ref}} = 0\) for all students in the reference group and \({\eta_i}^{\text{ref}} = -1\) for all students in the focal group. We now need to use a compensatory 2PL model where the slope on target ability \(a_{1j} = 1\) for all items (consistent with the Rasch model) and the slope on nuisance ability \(a_{2j} = 0.6\) for all items with DIF, and \(a_{2j} = 0\) otherwise. Again, 0.6 is the magnitude of DIF in logits. According to the two-dimensional compensatory 2PL model, the probability student \(i\) responding correctly to item \(j\) is
\[
\text{Pr}(y_{ij} = 1 | \theta_i, \eta_i, a_{1j}, a_{2j}, b_j) = \sigma(a_{1j}\theta_i + a_{2j}\eta_i + b_j).
\]
This translation between views makes explicit that nearly all DIF simulation studies have, perhaps suboptimally, examined the unrealistic scenario in which there is no variation in nuisance ability across students in the same group.

@@

\hypertarget{assumptionless-identification-methods}{%
\section{Assumptionless identification methods}\label{assumptionless-identification-methods}}

Psychometricians have long been in search of the perfect DIF detection method. It has become well known that the IRT-based likelihood ratio test (LRT) is the best way to test for DIF, which has been well demonstrated (Meade and Wright \protect\hyperlink{ref-meade2012solving}{2012}). As a result, we avoid methods like the Mantel-Haenszel procedure (Holland and Thayer \protect\hyperlink{ref-holland1986differential}{1986}), which muddies the waters by moving away from the IRT framework, and has been shown to perform no better --- and in some cases worse --- than IRT-based methods (Swaminathan and Rogers \protect\hyperlink{ref-swaminathan1990detecting}{1990}).

The unsolved and interesting problem is how to link groups in the common circumstance in which no a priori assumptions --- about either relative group ability or which items contain bias --- can be made. That is, we're interested in assumptionless identification methods (hereafter referred to as \enquote{AI methods}) so that the LRTs can be used to test items for DIF. For simplicity, we focus on the Rasch model to isolate the fundamental issues in DIF detection. And, in search of a coherent framework, we sometimes edit names of existing methods. We recognize that others have done the same (e.g.~Kopf, Zeileis, and Strobl \protect\hyperlink{ref-kopf2015framework}{2015}), and that we risk contributing to the proliferation of names.

In this section, we summarize existing AI methods and propose extensions of those methods. We use a simple, one-run simulation to demonstrate the methods: 10,000 reference group students and 10,000 focal group students taking an eight-item test. Target ability is simulated \({\theta_i}^{\text{ref}} \sim N(0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(-1,1)\) for students in the focal group. Nuisance ability is set to \({\eta_i}^{\text{ref}} = 0\) and \({\eta_i}^{\text{ref}} = -1\). The slope on target ability is set to \(a_{1j} = 1\) for all items. The slope on nuisance ability is set to \(a_{2j} = 0.5\) for the last three items (the items with DIF) and \(a_{2j} = 0\) otherwise. From the DIF-as-item-property view, there is no nuisance ability. Instead, \({b_j}^{\text{foc}} = {b_j}^{\text{ref}}\) for the first five items and \({b_j}^{\text{foc}} = {b_j}^{\text{ref}} - 0.5\) for the last three items.

In general, we denote the mean ability of the reference and focal group as \(\mu^\text{ref}\) and \(\mu^\text{foc}\), respectively. We follow the common and inconsequential practice of identifying the scale by setting \(\mu^\text{ref} = 0\). The fundamental challenge is to use an AI method to identify the model, which allows for the disentangling \(\hat\mu^\text{foc}\) from estimating \(\hat{b_j}^{\text{foc}} - \hat{b_j}^{\text{ref}}\). The most common AI method is to select a group of anchor items that are assumed to be DIF-free. These anchor items identify the model, thereby allowing for the estimation of \(\hat\mu^\text{foc}\), and the remaining items can be tested for DIF using an LRT.

Computing is done in R (R Core Team \protect\hyperlink{ref-rcore}{2019}), model fitting in the R package mirt (Chalmers \protect\hyperlink{ref-chalmers2012mirt}{2012}), and data wrangling/visualization in the set of R packages known as the tidyverse (Wickham \protect\hyperlink{ref-tidy}{2017}). Code is available at TO DO.

\hypertarget{anchoritems}{%
\subsection{Anchor items}\label{anchoritems}}

@@

Meade and Wright (\protect\hyperlink{ref-meade2012solving}{2012}) compared the most commonly used AI methods and unequivocally recommended the all-others-as-anchors (AOAA) method. AOAA tests each item for DIF one at a time using all of the other items as anchors. For example, when testing the first item for DIF, all of the other items are used as anchors. This is done using an LRT that compares the baseline model, where all item parameters are fixed across groups, to the flexible model, where the parameters of the tested item are freed across groups (Thissen, Steinberg, and Wainer \protect\hyperlink{ref-thissen1993detection}{1993}). Then, when testing the second item for DIF, once again all of the other items (including the first item) are used as anchors, and so on. The items for which the flexible model outperforms the baseline model (typically based on a \(\chi^2\) test) are identified as having DIF, and the rest of the items become anchor items. Note that AOAA is implemented in the mirt R package, and is called by passing scheme = \enquote{drop} to the DIF function (drop refers to dropping a single constraint when moving from the constrained to the flexible model).

Edelen et al. (\protect\hyperlink{ref-edelen2006identification}{2006}) used AOAA to look for DIF between the English and Spanish versions of the 21-item Mini-Mental State Examination and found that 10 of the 21 items exhibited DIF. How can they be sure that its those 10 items and not the other 11 items with DIF? They cannot be. Implicit in the use of AOAA is the assumption that all items not being tested do not exhibit DIF, which is, of course, impossible. More practically, it is thought that AOAA will perform well if a small minority of items have DIF or the DIF is balanced such that some items are biased against the focal group, while others are biased against the reference group. Undesirably, most applications of AI methods and many simulation studies do not make explicit the assumptions of the AI method (Strobl et al. \protect\hyperlink{ref-strobl2018anchor}{2018}). In this way, psychometricians might benefit from adopting economists' habit of explicitly stating assumptions and debating their plausibility.

@@

One way to attempt to counter artificial DIF is with purification through iterative anchor selection. For example, Drasgow (\protect\hyperlink{ref-drasgow1987study}{1987}) started with AOAA, removed items displaying DIF from the anchor set, then repeated the process iteratively --- with items that have been removed from the anchor set allowed to have free parameters across groups in both the baseline and flexible model --- until no more items tested positively. Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) named this technique Iterative-backward-AOAA with \enquote{backward} (as in reverse, not incorrect) referring to beginning with the assumption that all items are DIF-free. We find it clearer to refer to this method as all-others-as-anchors-all-significiant (AOAA-AS). Appending all-significant indicates that anchor selection is made iteratively with all items that test positive for DIF being removed from the anchor set. Note that AOAA-AS is implemented in the mirt R package, and is called by passing scheme = \enquote{drop\_sequential} to the DIF function.

AOAA-AS might seem like an improvement, but it doesn't solve a fundamental problem of AOAA: What does one do when all items test positive for DIF? With a sufficient sample size and at least one item with DIF, this will necessarily be the case. In our thought experiment, we get the same result with AOAA-AS as we did with AOAA: All items test positive for DIF, and there are no anchor items. Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) encountered precisely this problem in their simulation study and chose to select a single anchor item randomly. Woods (\protect\hyperlink{ref-woods2009empirical}{2009}) suggested a more straightforward, one-step method which uses AOAA and selects the, say, four items that exhibit the least amount of DIF. It's unclear if one should proceed if those four items display DIF.

@@

We propose an extension of these methods, all-others-as-anchors-one-at-a-time (AOAA-OAT), which, to our knowledge (and surprise), has not previously been explicitly proposed. AOAA-OAT is inspired by Hagquist and Andrich (\protect\hyperlink{ref-hagquist2017recent}{2017}), who, in general, assert that \enquote{items showing DIF initially should not be resolved simultaneously but sequentially} {[}p.~6{]}. Like AOAA-AS, AOAA-OAT starts with AOAA, but only the single item exhibiting the most DIF, based on the \(\chi^2\) test statistic, is removed from the anchor set. The process continues iteratively until no new items display DIF. AOAA-OAT and AOAA-AS are similar in that they are both iterative; the difference is that AOAA-OAT takes the more conservative approach of removing only one item in each iteration as opposed to all items that test positive for DIF. As a result, we believe that AOAA-OAT is less likely to be \enquote{fooled} by artifical DIF. Note that AOAA-OAT is not currently implemented in the R package mirt.

@@

Table \ref{table:themethods} summarizes the three all-others-as-anchors methods. It's useful to remember that AOAA is not an iterative procedure. The methods with a hyphen, AOAA-OAT and AAOA-AS, are iterative procedures with AOAA-OAT being our new, more conservative method. In our one-run simulation, AOAA and AOAA-AS both failed to select any anchor items, leaving the model unidentified. On the other hand, AOAA-OAT worked perfectly, selecting exactly the first five items as anchor items.

\begin{table}[H]
\caption{Summary of the three all-others-as-anchors assumptionless identification methods}
\centering
\begin{tabular}{|p{2.5cm}|p{6cm}|p{4cm}|}
\toprule

@@

@@

Bechger and Maris (\protect\hyperlink{ref-bechger2015statistical}{2015}) proposed selecting anchor items by identifying clusters of items that function similarly and then choosing one of those clusters to be the \enquote{anchor cluster.} They pointed out that one way around the unidentifiability issue is to consider only relative item parameters. For each group, the relative easinesses for each pair of items can be stored in the matrix \(\mathbf{R}^{\text{ref}}\) with entries \({R_{xy}}^{\text{ref}} = {b_x}^{\text{ref}} - {b_y}^{\text{ref}}\). The ultimate matrix of interest is \(\Delta \mathbf{R} \equiv \mathbf{R}^{\text{ref}}-\mathbf{R}^{\text{foc}}\) which is the \enquote{differences between groups in the pairwise differences in (easiness) between items} {[}p.~323{]}.

The general idea of identifying clusters of items is intriguing. However, their approach is needlessly complicated, and they did not describe a process for moving from \(\Delta \mathbf{R}\) to an anchor cluster. Pohl, Stets, and Carstensen (\protect\hyperlink{ref-pohl2017cluster}{2017}) extended their work by proposing one such process. \(\Delta \mathbf{R}\) is skew-symmetric and of rank 1, which means that all information is contained in a single row or column. Accordingly, they use k-means clustering on just the first column of \(\Delta \mathbf{R}\) where the number of clusters, k, is chosen by minimizing BIC. They suggest using a combination of cluster size, cluster homogeneity, and cluster parameter precision to choose which of the clusters is the anchor cluster. Unfortunately, in their simulation study, they find that BIC identifies only a single cluster, so they end up using all of the items as anchors.

We propose a new cluster-based approach, which we call \enquote{equal means clustering} (EMC). Instead of working with an arbitrary column from \(\Delta \mathbf{R}\), we work with the vector \(\tilde{\mathbf{d}}\) of differences in item easiness, \(\tilde{d_j} = \tilde{b_j}^\text{ref} - \tilde{b_j}^\text{foc}\), where the model is identified by setting \(\mu^\text{foc} = 0\) (recall \(\mu^\text{ref}\) is always set to \(0\), thus the name \enquote{equal means clustering}). As a result, all differences in performance --- either from group ability differences or DIF --- manifesting in the item easiness difference parameter \(\tilde{d_j}\). The tilde is used to denote parameters estimated with \(\mu^\text{foc}\) arbitrarily set to \(0\).
